Intrusion Detection with \\ Genetic Algorithms and Fuzzy Logic

### 2013-11-14 paper.tex
Made some changes based on feedback from Alex.

\title{Intrusion Detection with Genetic Algorithms and Fuzzy Logic}

@@

\maketitle
\begin{abstract}
This paper describes two different ways of training an intrusion detection system about possible attacks to a system: genetic algorithms and fuzzy logic. I will explain how genetic algorithms are used in intrusion detection systems and compare the results to the winning entry of the KDD99 Classifier Learning Contest. I will then explain how fuzzy genetic algorithms are used and compare those results with a decision tree algorithm.
\end{abstract}

@@

In Section~\ref{sec:background} I will give some background on intrusion detection. I will explain the main types of networking attacks in Section~\ref{sec:typesAttacks}, and ways of detecting attacks in Section~\ref{sec:detectionMeth}. In Section~\ref{sec:dataSets} I will describe two data sets that are used in intrusion detection. I will explain what \emph{rules} are in Section~\ref{sec:rules}, and then give information on fuzzy logic and genetic algorithms in Sections~\ref{sec:fuzzyLogic} and~\ref{sec:GA}. After that, I will explain the measures that are used in determining the accuracy of an intrusion detection system in Section~\ref{sec:PosNeg}. In Section~\ref{sec:genAlgImp} I will explain the use of a genetic algorithm in an intrusion detection system, and in Section~\ref{sec:fuzGenAlgImp} I will explain the use of a fuzzy genetic algorithm. Section~\ref{sec:conclusion} has some conclusions about the use of genetic algorithms and fuzzy genetic algorithms in intrusion detection systems.

@@

\subsection{Data Sets}
\label{sec:dataSets}
Two different data sets are used in this paper: KDD99 and RLD09. The KDD99 data set is a benchmark data set that was generated by simulating a military network environment in 1999~\cite{6559603}. It has long been a standard data set for intrusion detection. The data in the set is classified as normal or attack activity. KDD99 uses 41 \emph{features}, which are properties of a \emph{record} (either an attack or normal activity), that are used to describe the activity and help to distinguish normal connections from attacks.

In research conducted by P. Jongsuebsuk, N. Wattanapongsakorn, and
C. Charnsripinyo~\cite{6496342, 6559603}, 8 out of the 41 KDD99 features~\cite{KDD99} were used in their system:
\begin{enumerate}
  \item duration: the length of the normal or attack activity in seconds.
  \item src\_bytes: the number of bytes sent from source to destination. Source is the user who may or may not be an attacker, and destination is the server being potentially attacked.
  \item num\_failed\_logins: the number of failed login attempts.
  \item root\_shell: returns 1 if root shell is obtained, which means that the user is able to log in as root. This gives them the ability to do things like add accounts and change user passwords. If root shell is not obtained, 0 is returned.
  \item num\_access\_files: the number of operations on access control files. Access control files specify which users are granted access to objects and what operations are allowed on the objects. An example would be (Alice, delete), which would give Alice permission to delete the file.~\cite{accessControl}
  \item srv\_count: the number of connections to the same service as the current connection in the past two seconds.
  \item serror\_rate: the percentage of connections that have ``SYN" errors. When a client attempts to connect to a server, it first sends a SYN (synchronize) message to the server. The server then acknowledges the request by sending a SYN-ACK to the client. The connection is established when the client sends an ACK back to the server. A SYN error is a failure that happens early in this process.~\cite{TCP}
  \item same\_srv\_rate: the percentage of connections to the same service.
\end{enumerate}

The KDD99 data set is 14 years old, and newer attack types are not included in it because of its age. Because of this, P. Jongsuebsuk, N. Wattanapongsakorn, and
C. Charnsripinyo~\cite{6496342, 6559603} created their own data set, RLD09, to use in their experiments. To create the data set, the authors captured network data from the Computer Engineering Department at King Mongkut's University of Technology Thonburi, in Bangkok, Thailand. The data has around ten million preprocessed data \emph{packets}. A packet is a unit of data that is carried throughout a network~\cite{networkPacket}. RLD09 has 17 different types of attacks that can be divided into denial of service attacks and probe attacks. It also has normal network activity. RLD09 uses 12 features, which include the number of packets, source ports, and destination ports.

\subsection{Rules}
\label{sec:rules}
Rules are a way in which elements of one set are separated into different sets, or classes, in order to differentiate between normal connections and attacks. Rules are represented by if-then statements in the following format: If <\emph{condition}> then <\emph{action}>.~\cite{bapuji2012soft} Rules can specify the details of a packet such as the IP address, port number and protocal. If a packet matches any of the rules in the intrusion detection system then the system will take appropriate action, which may include stopping the connection or logging off the system.~\cite{DBLP:journals/corr/abs-1204-6416}

\subsection{Fuzzy Logic}
\label{sec:fuzzyLogic}
Attacks on systems do not always have a fixed pattern, so fuzzy logic is used to detect patterns that have a behavior that is between normal and unusual. A fuzzy logic rule has the following format: If <\emph{condition}> then <\emph{consequence}>, where \emph{condition} is a fuzzy variable, and \emph{consequence} is a fuzzy set. For example: if the number of packets with the same destination address is high, then the pattern is unusual. To determine what is considered high, the values of the packets are divided into fuzzy sets. In Figure~\ref{fig:triangleFigure} there are three sets: low (region A), medium (region B), and high (region C). 
\begin{figure}
\centering
\psfig{file=triangleFigure.JPG}
\caption{Fuzzy shape with three sets: low (A), medium (B), and high (C)~\cite{DBLP:journals/corr/abs-1204-6416}}
\label{fig:triangleFigure}
\end{figure}
The x-axis are the values in the fuzzy set (number of packets), and the y-axis is the membership function, which determines the region. For example, if the number of packets with the same destination address is 15 then this will be considered as low for a degree of 0.4, but will be considered medium or high for other degrees. If the number of packets is determined to be high, then the connection will be aborted.~\cite{DBLP:journals/corr/abs-1204-6416}

\subsection{Genetic Algorithm}
\label{sec:GA}
Genetic algorithms are a search technique used to find solutions to problems. \emph{Mutation}, \emph{selection}, and \emph{crossover} are used to evolve and improve rules.

@@

The \emph{detection rate} (DR) of an intrusion detection system is the number of true positives divided by the total number of intrusions that happen.~\cite{ids}

\section{Genetic Algorithm Implementation}
\label{sec:genAlgImp}

\subsection{Algorithm Overview}
In research conducted by M. S. Hoque, M. A. Mukit, and M. A. N. Bikas~\cite{DBLP:journals/corr/abs-1204-1336}, a genetic algorithm was used to make their intrusion detection system. The system is divided into two phases: a precalculation phase and a detection phase. In the precalculation phase, a set of chromosomes are created using training data. Then this set of chromosomes is used in the detection phase for comparison. Algorithm~\ref{alg:precalc} is used in the precalculation phase.

@@

In the detection phase, a population is created for test data. Selection, crossover, and mutation occur, and then the type of the data (whether it is an attack or normal behavior) is predicted. The set of chromosomes that was created in the precalculation phase is used in the detection phase to find the fitness of each chromosome in the population. Algorithm~\ref{alg:detection} is used in the detection phase.

@@

    chromosome}
  \ENDFOR
  \STATE{Remove some chromosomes with worse fitness}
  \STATE{Apply crossover to the selected pair of chromosomes of the population}
  \STATE{Apply mutation to each chromosome of the

@@

\subsection{Experimental Design and Results}
The authors of~\cite{DBLP:journals/corr/abs-1204-1336} used the KDD99 data set. They used only the numerical features of the KDD99 data set (34 out of 41 total features). For the training data set, the KDD99 10\% version file was used, and for the test data set, the KDD99 corrected version file~\cite{KDD99} was used. The training set has a total of 494,021 records, and 396,741 of them are attacks. The test set has a total of 311,029 records, and 250,436 of them are attacks. Table~\ref{tab:numberOfRecords} shows the number of records of normal and attack activity in the training and test sets.

@@

In the detection phase, an initial population is created and then is compared with each chromosome that was created in the precalculation phase. Crossover and mutation happen in the population in order to create a new population. The results from running the genetic algorithm are shown in Tables~\ref{tab:genAlgResults} and~\ref{tab:genAlgResults2}. The detection rate was 94.9\%. The false positive rate was 30.5\%. The false negative rate was 5\%. The true negative rate was 69.5\%.

The authors of~\cite{DBLP:journals/corr/abs-1204-1336} compared their results with the winning entry of the KDD99 Classifier Learning Contest~\cite{KDD99Contest}, and found that they had a better detection rate for denial of service and user to root attacks than the winning entry.

\section{Fuzzy Genetic Algorithm Implementation}
\label{sec:fuzGenAlgImp}
The focus of~\cite{6496342, 6559603} is on detecting new or unknown types of attacks in a network. The intrusion detection system used is able to identify normal network activity as well as attacks using a fuzzy genetic algorithm. This kind of algorithm is able to learn new attacks, and has a high detection rate.

\subsection{Fuzzy Algorithm}
In order to measure the probability that a record is an attack, a trapezoidal shape was used in the algorithm. This is shown in Figure~\ref{fig:trapFigure}. The trapezoidal shape has four parameters: $a, b, c$, and $d$. Algorithm~\ref{alg:fuzAlg} calculates the probability of a record being an attack.

\begin{figure}
\centering
\psfig{file=trapFigure.jpg}
\caption{Trapezoidal shape with 4 parameters where $a\leq b\leq c\leq d$}
\label{fig:trapFigure}
\end{figure}

@@

@@

  \STATE{compare the predicted result with actual result}
  \STATE{find $A$, $B$, $\alpha$, and $\beta$}
  \STATE{// $A$ is total number of attack records. $B$ is total number of normal records. $\alpha$ is total number of attack records correctly identified as attack. $\beta$ is total number of normal records incorrectly classified as attack.}
\ENDFOR
\STATE{calculate fitness}
\STATE{//create next generation}
\STATE{preserve\_best()}
\STATE{crossover()}
\STATE{mutation() // mutation rate~=~0.30}
\end{algorithmic}
\end{algorithm}

@@

In this implementation, a population size of 10 was used for each generation. An individual in the population represents a possible detection rule. The two best individuals from a present generation are preserved for the next generation. The other individuals in the new generation come from mutation and single-point crossover.

\subsection{Experimental Design and Results}

\subsubsection{Experiments using Only RLD09}
The experiments that the authors of~\cite{6496342, 6559603} performed used a total of 16,000 records of normal activity and 10,500 records of attack activity. Of the attack records, 4,000 of them were denial of service attacks and 6,500 were probe attacks.

@@

\begin{table*}
\center
\caption{Results from Experiment 1}
\vspace{0.20cm}
\begin{tabular}{ccccccc}
 & Attack & Normal & Total Records & FP (\%) & FN (\%) & DR (\%)\\
DoS Training & 1499 & 8501 & 10000 & 1.46 & 47.50 & 91.64\\
Probe Training & 2496 & 7504 & 10000 & 1.83 & 15.38 & 94.79\\
Testing & 10500 & 16000 & 26500 & 1.13 & 4.10 & 97.92\\
\end{tabular}
\label{tab:fuzGenExp1}
\center
\end{table*}

In the second experiment, seven tests were run. For each test case there were 13 attack types plus normal activity that were in the training data set. Three attack types were used for the unknown testing data set. For example, test case 1 used the training data set that does not have Advance Port Scan, Ack Scan, and Xmas Tree, which are all probe attacks. These three attacks were then used for the testing data set. Table~\ref{tab:fuzGenExp2} shows some of the results from the fuzzy genetic algorithm and a decision tree algorithm, which is another common way of addressing these problems. For further information on decision trees, see~\cite{decisionTree}. When compared with the decision tree algorithm, the fuzzy genetic algorithm has a higher detection rate in all cases except 3 and 5. It can be seen that in cases 2 and 4 the decision tree has low detection rates, while the fuzzy genetic algorithm has much higher detection rates.

\begin{table}
\caption{Unknown Attack Experiment}
\vspace{0.20cm}
\begin{tabular}{llll}
Test & Unknown & Decision & Fuzzy\\
Case & Attacks & Tree DR (\%)  & Genetic DR (\%)\\ \hline

@@

\subsubsection{Experiments using Both RLD09 and KDD99}
The authors of~\cite{6496342, 6559603} also ran experiments that used and compared the RLD09 data set with the KDD99 data set. They used the KDD99 10\% version file~\cite{KDD99} for both the training dataset and testing dataset.

The authors of~\cite{6496342, 6559603} first trained and tested the fuzzy genetic algorithm (Algorithm~\ref{alg:fuzGenAlg}) with the KDD99 data set. There were 6 different types of denial of service attacks and 4 different types of probe attacks. The detection rate of the KDD99 data set was 98.72\%. Then 26,500 records of the RLD09 data set were used as the training set. The detection rate was 97.97\%. The results of this experiment are shown in Table~\ref{tab:bothSetsResults}.

@@

The next experiment was the use of the KDD99 training set with the fuzzy genetifc algorithm to separate the data into two classes. Then each specific attack type was extracted and combined with normal activity. Ten tests were run, and Table~\ref{tab:kddAttacks} shows the accuracy of detecting some of the cases. The results showed that the detection rate of most of the cases were greater than 93\%. The results also showed that the behavior of the attacks were highly distinctive from normal activity. There were only two cases that had low detection rates, one of which is case 1 in Table~\ref{tab:kddAttacks}.

@@

The final experiment that was run used only the RLD09 data set with the fuzzy genetic algorithm. 17 tests were run, and Table~\ref{tab:rldAttacks} shows the accuracy of detecting some of the cases. The results showed that the detection rate of a majority of the cases were greater than 97.5\%. Again, there were only two test cases that had low detection rates, one of which is case 2 in Table~\ref{tab:rldAttacks}.

>>>>

\title{Intrusion Detection with \\ Genetic Algorithms and Fuzzy Logic}

@@

\maketitle
\begin{abstract}
This paper describes two different ways of training an intrusion detection system about possible attacks to a system: genetic algorithms and fuzzy logic. I will describe how genetic algorithms are used in intrusion detection systems and compare the results to the winning entry of the KDD99 Classifier Learning Contest. I will then describe how fuzzy genetic algorithms are used and compare those results with a decision tree algorithm.
\end{abstract}

@@

In Section~\ref{sec:background} I will give some background on intrusion detection. I will describe the main types of networking attacks in Section~\ref{sec:typesAttacks}, and ways of detecting attacks in Section~\ref{sec:detectionMeth}. In Section~\ref{sec:dataSets} I will describe two data sets that are commonly used in intrusion detection. I will explain how \emph{rules} are used to differentiate between normal connections and attacks in Section~\ref{sec:rules}, and then give information on genetic algorithms in Section~\ref{sec:GA}. After that, I will describe the measures that are used in determining the accuracy of an intrusion detection system in Section~\ref{sec:PosNeg}. In Section~\ref{sec:genAlgImp} I will describe the use of a genetic algorithm in an intrusion detection system, and in Section~\ref{sec:fuzGenAlgImp} I will describe the use of a fuzzy genetic algorithm. Section~\ref{sec:conclusion} has some conclusions about the use of genetic algorithms and fuzzy genetic algorithms in intrusion detection systems.

@@

\subsection{Data Sets}
\label{sec:dataSets}
Two different data sets are used in this paper to evaluate the performance of intrusion detection systems: KDD99 and RLD09. The KDD99 data set is a benchmark data set that was generated by simulating a military network environment in 1999. It has long been a standard data set for intrusion detection. The data was processed into five million records. A record is a sequence of TCP packets, between which data flows to and from a source IP address to a target IP address. The data in the set is classified as either normal or attack activity. KDD99 uses 41 \emph{features}, which are properties of a record that are used to describe the activity and help to distinguish normal connections from attacks.~\cite{KDD99}

In research conducted by P. Jongsuebsuk,

N. Wattanapongsakorn, and
C. Charnsripinyo~\cite{6496342, 6559603}, which will be discussed in Section~\ref{sec:fuzGenAlgImp}, eight out of the 41 KDD99 features~\cite{KDD99} were used. They decided on these 8 features based on research done in~\cite{Fries:2008:FAN:1388969.1389037}. The features are:
\begin{enumerate}
  \item duration: the length of the normal or attack activity in seconds.
  \item src\_bytes: the number of bytes sent from source to destination. Source is the user who may or may not be an attacker, and destination is the server being potentially attacked.
  \item num\_failed\_logins: the number of failed login attempts.
  \item root\_shell: returns 1 if root shell is obtained, which means that the user is able to log in as root. This gives them the ability to do things like add accounts and change user passwords. If root shell is not obtained, 0 is returned.
  \item num\_access\_files: the number of operations on access control files. Access control files specify which users are granted access to objects and what operations are allowed on the objects. An example would be (Alice, delete), which would give Alice permission to delete the file.~\cite{accessControl}
  \item srv\_count: the number of connections to the same service as the current connection in the past two seconds.
  \item serror\_rate: the percentage of connections that have ``SYN" errors. When a client attempts to connect to a server, it first sends a SYN (synchronize) message to the server. The server then acknowledges the request by sending a SYN-ACK to the client. The connection is established when the client sends an ACK back to the server. A SYN error is a failure that happens early in this process.~\cite{TCP}
  \item same\_srv\_rate: the percentage of connections to the same service.
\end{enumerate}

The KDD99 data set is 14 years old, and newer attack types are not included in it because of its age. Because of this, P. Jongsuebsuk, N. Wattanapongsakorn, and
C. Charnsripinyo~\cite{6496342, 6559603} created their own data set, RLD09, to use in their experiments. To create the data set, the authors captured network data from a university in Bangkok, Thailand. RLD09 has 17 different types of attacks that can be divided into denial of service attacks and probe attacks. It also has normal network activity. RLD09 uses 12 features, which include the number of packets, source ports, and destination ports. For further information, see~\cite{6496342, 6559603}. RLD09 only has 12 features because the number of features affect the speed and memory consumption of the intrusion detection system.

\subsection{Rules}
\label{sec:rules}
A commonly used approach for detecting intrusions and to differentiate between normal connections and attacks is to use rules. Rules are represented by if-then statements in the following format: If (\emph{condition}) then (\emph{consequence}). The condition part of the rule is composed of one or more features, and the consequence of the rule says if it is an intrusion or not. For example: if (\emph{duration} = 1, and \emph{src\_bytes} = 0, and \emph{serror\_rate} = 50) then \emph{intrusion}.~\cite{Fries:2008:FAN:1388969.1389037}

\subsection{Genetic Algorithm}
\label{sec:GA}
Genetic algorithms are a search technique used to find solutions to problems. Operations analogous to biological \emph{mutation}, \emph{selection}, and \emph{crossover} are used to evolve and improve rules.

@@

The \emph{detection rate} (DR) of an intrusion detection system is the number of intrusions detected by the system divided by the total number of intrusions.~\cite{ids}

\section{Using Genetic Algorithms}
\label{sec:genAlgImp}

\subsection{Algorithm Overview}
In research conducted by M. S. Hoque, M. A. Mukit, and M. A. N. Bikas~\cite{DBLP:journals/corr/abs-1204-1336}, a genetic algorithm was used to make their intrusion detection system. The system is divided into two phases: a precalculation phase and a detection phase. In the precalculation phase, a set of chromosomes are created using training data. Then this set of chromosomes is used in the detection phase for comparison. Algorithm~\ref{alg:precalc} is used in the precalculation phase. It initializes chromosomes for comparison. It takes the network data as an input and outputs a set of chromosomes.

@@

In the detection phase, the precalculated set of chromosomes is used to find the fitness of each chromosome in the population. Selection, crossover, and mutation occur, and then the type of the data (whether it is an attack or normal behavior) is predicted. Algorithm~\ref{alg:detection} is used in the detection phase. It takes the precalculated set of chromosomes as an input and outputs the type of data.

@@

    chromosome}
  \ENDFOR
  \STATE{Remove some chromosomes with worse fitness}
  \STATE{Apply crossover to the selected pair of chromosomes of the population // crossover rate~=~0.15}
  \STATE{Apply mutation to each chromosome of the

@@

\subsection{Experimental Design and Results}
The authors of~\cite{DBLP:journals/corr/abs-1204-1336} used the KDD99 data set. They used only the numerical features of the KDD99 data set (34 out of 41 total features). For the training data set, the KDD99 10\% version file was used. This is a 10\% subset of the full KDD99 data set. For the test data set, the KDD99 corrected version file was used. This file corrects various typos that were in other files.~\cite{KDD99} The training set has a total of 494,021 records, and 396,741 of them are attacks. The test set has a total of 311,029 records, and 250,436 of them are attacks. Table~\ref{tab:numberOfRecords} shows the number of records of normal and attack activity in the training and test sets.

@@

In the detection phase, an initial population is created and then is compared with each chromosome that was created in the precalculation phase. Crossover and mutation happen in the detection population in order to create a new population. The results from running the genetic algorithm are shown in Tables~\ref{tab:genAlgResults} and~\ref{tab:genAlgResults2}. The detection rate was 94.9\%. The false positive rate was 30.5\%. The false negative rate was 5\%. The true negative rate was 69.5\%.

The authors of~\cite{DBLP:journals/corr/abs-1204-1336} compared their results with the winning entry of the KDD99 Classifier Learning Contest~\cite{KDD99Contest}. The winning entry used a decision tree algorithm, which is another common algorithm that is used. For further information on decision trees, see~\cite{decisionTree}. The authors of~\cite{DBLP:journals/corr/abs-1204-1336} found that they had a better detection rate for denial of service and user to root attacks than the winning entry. For the winning entry, the detection rate of denial of service was 97.1\%, and for user to root it was 13.2\%. In~\cite{DBLP:journals/corr/abs-1204-1336} the detection rate of denial of service was 99.4\% and for user to root it was 18.9\%.

\section{Using Fuzzy Genetic Algorithms}
\label{sec:fuzGenAlgImp}
The focus of~\cite{6496342, 6559603} is on detecting new or unknown types of attacks in a network. The intrusion detection system used is able to identify normal network activity as well as attacks using a fuzzy genetic algorithm. This kind of algorithm is able to learn new attacks, and has a high detection rate.

\subsection{Fuzzy Logic Rules and Algorithm}
Attacks on systems do not always have a fixed pattern, so fuzzy logic is used to detect patterns that have a behavior that is between normal and unusual. Fuzzy logic rules are similar to the rules described in Section~\ref{sec:rules}, except that \emph{consequence} is a certainty factor. For example, if (\emph{duration} = 6) then (\emph{probability of it being an attack is 50\%}).

In order to measure the probability that a record is an attack, a trapezoidal shape was used in~\cite{6496342, 6559603}. This is shown in Figure~\ref{fig:trapFigure}. The trapezoidal shape has four parameters: $a, b, c$, and $d$. Algorithm~\ref{alg:fuzAlg} calculates the probability of a record being an attack.

\begin{figure}
\centering
\psfig{file=TrapezoidFigure.pdf, width=3.3in}
\caption{Used to measure the probability that a record is an attack}
\label{fig:trapFigure}
\end{figure}

@@

An example of using the fuzzy shape and algorithm is: suppose that the feature is duration, and suppose it is 6 seconds. Suppose that $a$ = 1, $b$ = 3, $c$ = 5, $d$ = 7. Then the probability of an attack is equal to

\begin{equation*}
\frac{d-\textrm{data}}{d-c} = \frac{7-6}{7-5} = 0.5.
\end{equation*}

@@

  \STATE{compare the predicted result with actual result}
  \STATE{find $A$, $B$, $\alpha$, and $\beta$}
  \STATE{// $A$ is total number of attack records. $B$ is total number of normal records. $\alpha$ is total number of attack records correctly identified as attack. $\beta$ is total number of normal records incorrectly classified as attack.}
\ENDFOR
\STATE{calculate fitness}
\STATE{//create next generation}
\STATE{preserve\_best()}
\STATE{crossover() // single-point crossover}
\STATE{mutation() // mutation rate~=~0.30}
\end{algorithmic}
\end{algorithm}

@@

A population size of 10 was used for each generation. An individual in the population represents a possible detection rule. The two best individuals from a present generation are preserved for the next generation. The other individuals in the new generation come from mutation and single-point crossover.

\subsection{Experimental Design and Results}
A variety of experiments were run in~\cite{6496342, 6559603}. Two experiments used just RLD09, and three experiments used KDD99 and RLD09 together.

\subsubsection{Experiments using Only RLD09}
The experiments using only RLD09 that the authors of~\cite{6496342, 6559603} performed used a total of 16,000 records of normal activity and 10,500 records of attack activity. Of the attack records, 4,000 of them were denial of service attacks and 6,500 were probe attacks.

@@

\begin{table*}
\center
\caption{Results from Experiment 1, using only RLD09}
\vspace{0.20cm}
\begin{tabular}{ccccccc}
 & Attack & Normal & Total Records & FP (\%) & FN (\%) & DR (\%)\\
DoS Training & 1499 & 8501 & 10000 & 1.46 & 47.50 & 91.64\\
Probe Training & 2496 & 7504 & 10000 & 1.83 & 15.38 & 94.79\\
Testing & 10500 & 16000 & 26500 & 1.13 & 4.10 & 97.92\\
\end{tabular}
\label{tab:fuzGenExp1}
\center
\end{table*}

In the second experiment,~\cite{6496342, 6559603} pulled some types of attacks out of the training set and kept them for unknown data testing. This was to test that the fuzzy genetic algorithm could detect unknown attacks. In this experiment, seven tests were run. For each test case there were 13 attack types plus normal activity that were in the training data set. Three attack types were used for the unknown testing data set. For example, test case 1 used the training data set that does not have Advance Port Scan, Ack Scan, and Xmas Tree, which are all probe attacks. These three attacks were then used for the testing data set. Table~\ref{tab:fuzGenExp2} shows some of the results from the fuzzy genetic algorithm and a decision tree algorithm, which is another common way of addressing these problems. For further information on decision trees, see~\cite{decisionTree}, and for further information on the types of denial of service and probe attacks, see~\cite{typesOfAttacks}. When compared with the decision tree algorithm, the fuzzy genetic algorithm has a higher detection rate in all cases except 3 and 5.

\begin{table}
\caption{Unknown Attack Experiment, using only RLD09}
\vspace{0.20cm}
\begin{tabular}{llll}
Test & Unknown & Decision & Fuzzy\\
Case & Attacks & Tree DR (\%)  & Genetic DR (\%)\\ \hline

@@

\subsubsection{Experiments using Both RLD09 and KDD99}
The authors of~\cite{6496342, 6559603} also ran experiments that used both the RLD09 data set and the KDD99 data set in order to compare how the fuzzy genetic algorithm would perform on both. They used the KDD99 10\% version file~\cite{KDD99} for both the training dataset and testing dataset.

The first experiment used the fuzzy genetic algorithm to classify normal activity and attacks from both data sets. The authors of~\cite{6496342, 6559603} first trained and tested the fuzzy genetic algorithm (Algorithm~\ref{alg:fuzGenAlg}) with the KDD99 data set. There were 6 different types of denial of service attacks and 4 different types of probe attacks. The detection rate of the KDD99 data set was 98.72\%. Then 26,500 records of the RLD09 data set were used as the training set. The detection rate was 97.97\%. The results of this experiment are shown in Table~\ref{tab:bothSetsResults}.

@@

The next experiment used the fuzzy genetic algorithm to classify types of attacks in the KDD99 data set. They used the KDD99 training set, with 158,597 records of denial of service attacks and 1,500 records of probe attacks. Ten tests were run, and Table~\ref{tab:kddAttacks} shows the accuracy of detecting some of the cases. The results showed that the detection rate of most of the cases were greater than 93\%. There were only two cases that had low detection rates, one of which is case 1 in Table~\ref{tab:kddAttacks}.

@@

The final experiment that was run used only the RLD09 data set with the fuzzy genetic algorithm to classify types of attacks. 17 tests were run, and Table~\ref{tab:rldAttacks} shows the accuracy of detecting some of the cases. The results showed that the detection rate of a majority of the cases were greater than 97.5\%. Again, there were only two test cases that had low detection rates, one of which is case 2 in Table~\ref{tab:rldAttacks}.
